Properties and Bayesian fitting of restricted Boltzmann machines

### 2018-09-27 writing/paper-rbm/paper.Rmd
grammar check

---
title: "Properties and Bayesian fitting of restricted Boltzmann machines"
author: 
  - name: Andee Kaplan
    shortname: Kaplan A.,
    numaddress: 1
    corresponding: true
  - name: Daniel Nordman
    shortname: D. Nordman, and
    corresponding: false
    numaddress: 2
  - name: Stephen Vardeman
    shortname: S. Vardeman
    numaddress: "2,3"
    corresponding: false
shortauthor: "KAPLAN \\textsc{et al}"
address:
  - numaddress: 1
    orgdiv: Department of Statistical Science
    orgname: Duke University
    orgstate: North Carolina
    orgcountry: USA
  - numaddress: 2
    orgdiv: Department of Statistics
    orgname: Iowa State University
    orgstate: Iowa
    orgcountry: USA
  - numaddress: 3
    orgdiv: Department of Industrial and Manufacturing Systems Engineering
    orgname: Iowa State University
    orgstate: Iowa
    orgcountry: USA
corresponding: true
corresponding_name: Andee Kaplan
corresponding_address: Department of Statistical Science, Duke University, P.O. Box 90251, Durham, NC 27708-0251
corresponding_email: andrea.kaplan@duke.edu
output: 
  rmarkdown::pdf_document:
    keep_tex: true
    template: ../resources/latex/wileynjd-ams_article.tex
    includes:
      in_header: ../resources/latex/header.tex
    toc: no
bibliography: [../resources/latex/refs_rbm.bib]
abstract: |
   A restricted Boltzmann machine (RBM) is an undirected graphical model constructed for discrete or continuous random variables, with two layers, one hidden and one visible, and no conditional dependency within a layer. In recent years, RBMs have risen to prominence due to their connection to deep learning. By treating a hidden layer of one RBM as the visible layer in a second RBM, a deep architecture can be created. RBMs are thought to thereby have the ability to encode very complex and rich structures in data, making them attractive for supervised learning. However, the generative behavior of RBMs is largely unexplored and typical fitting methodology does not easily allow for uncertainty quantification in addition to point estimates. In this paper, we discuss the relationship between RBM parameter specification in the binary case and model properties such as degeneracy, instability and uninterpretability. We also describe the associated difficulties that can arise with likelihood-based inference and further discuss the potential Bayes fitting of such (highly flexible) models, especially as Gibbs sampling (quasi-Bayes) methods are often advocated for the RBM model structure.
keywords: Degeneracy, Instability, Classification, Deep Learning, Graphical Models
---

@@

The data mining and machine learning communities have recently shown great interest in deep learning, specifically in stacked restricted Boltzmann machines (RBMs) [see @salakhutdinov2009deep; @salakhutdinov2012efficient; @srivastava2013modeling; @le2008representational for examples]. A RBM is a probabilistic undirected graphical model (for discrete or continuous random variables) with two layers, one hidden and one visible, with no conditional dependency within a layer [@smolensky1986information]. These models have reportedly been used with success in classification of images [@larochelle2008classification; @srivastava2012multimodal]. However, the model properties are largely unexplored in the literature and the commonly cited fitting methodology remains heuristic-based and relies on rough approximation [@hinton2006fast]. Additionally, the current fitting methodology does not easily allow for interval estimation of parameters in the model, and instead only provides for point estimation. In this paper, we provide steps toward a fuller understanding of the model class and its properties from the perspective of statistical model theory, and we then explore the possibility of a rigorous fitting methodology with natural uncertainty quantification. We find the RBM model class to be concerning in two fundamental ways.

First, the models can be unsatisfactory as conceptualizations of how data are generated. That is, recalling @fisher1922mathematical, the aim of a statistical model is to represent data in a compact way. Neyman and Box further state that a model should "provide an explanation of the mechanism underlying the observed phenomena" [@lehmann1990model; @box1967discrimination]. At issue, simulation from RBMs can often produce data lacking realistic variability so that such models may thereby fail to satisfactorily reflect an observed data generation process. Such behavior relates to model degeneracy (or near degeneracy), which is a statistical concern in that many data processes of interest are realistically not degenerate in their spectrum of potential outcomes. For example, when sampling data from a nearly degenerate RBM used to model an imaging process, only a small set of output possibilities receives nearly all probability, and thus a sample of images will all be copies of the same one image (or small number of images). An example of ten 4-pixel images simulated from a nearly degenerate RBM model is compared to ten 4-pixel images simulated from a non-degenerate RBM model in Figure \ref{fig:sample-models}. The degenerate model places almost all probability on one outcome, causing the image to be generated repeatedly, whereas the non-degenerate model shows more potentially realistic variation. \par

@@

In addition to such degeneracy, we find that RBMs can easily exhibit types of instability, related to how sensitive the model probabilities can be to small difference in data outcomes. In practice, this may be seen when a single pixel change in an image results in a wildly different classification in an image classification problem. Occurrences of such behavior have recently been documented in RBMs [@li2014biclustering], as well as other deep architectures [@szegedy2013intriguing; @nguyen2014deep]. We describe (related) potential issues of model instability, degeneracy and uninterpretability for the RBM class (which are related properties) and examine the presence of these in Section \ref{explorations-of-model-properties-through-simulation} through simulations of small, manageable examples.

@@

Furthermore, these MCMC algorithms involve updating potentially many latent variables (hiddens) which can critically influence convergence in MCMC-based likelihood methods. Applying basic statistical principles in fitting RBM models of tractable size, we compare three fully Bayesian techniques involving MCMC, which are computationally more accessible than direct maximum likelihood and also aim to avoid parts of a RBM parameter space that yield unattractive models. As might be expected, with greater computational complexity comes an increase in fitting accuracy, but at the cost of practical feasibility.

As a factor compounding the computational challenges, issues in model fitting traceable to model flexibility are potentially more concerning. For a RBM model with enough hidden variables, any distribution for the visibles can be approximated arbitrarily well [@le2008representational; @montufar2011refinements; and @montufar2011expressive]. However, for binary data, the empirical distribution of an observed training set of visible variables provides a highest likelihood benchmark, before any parametric model class is even introduced and applied to obtain a refinement of model fit. As a consequence, we find that any fully principled fitting method based on the likelihood for a RBM with enough hidden variables will seek to reproduce the (discrete) empirical distribution of a training set. This aspect can be undesirable, and perhaps even unexpected compared to most modeling scenarios, in that no "smoothed distribution" may result when fitting a RBM model of sufficient size with a rigorous likelihood-based method. We are therefore led to be skeptical that models that involve these structures (like deep Boltzmann machines) can achieve useful prediction or inference in a principled way without intentionally limiting the flexibility of the fitted model.

@@

While applications of the RBM have been claimed to produce some successes in classification problems, it is unclear if the model class allows one to go beyond fitting to other statistical matters of importance in using the models, such as quantification of the uncertainty of estimation and the formulation of predictive distributions. These are closely tied to the probability properties of the RBM model for explaining data generation and our exposition intends to contribute to a better understanding in this direction.

@@

A restricted Boltzmann machine (RBM) is an undirected graphical model specified for discrete or continuous random variables, binary variables being most commonly considered. In this paper, we consider the binary case for concreteness. A RBM architecture has two layers, hidden ($\mathcal{H}$) and visible ($\mathcal{V}$), with no dependency connections within a layer. An example of this structure is in Figure \ref{fig:rbm} with the hidden nodes indicated by gray circles and the visible nodes indicated by white circles.
\begin{figure}
  \centering
  \resizebox{\linewidth}{!}{\input{../resources/images/tikz/rbm.tikz}}
  \caption{An example restricted Boltzmann machine (RBM), consisting of two layers, one hidden ($\mathcal{H}$) and one visible ($\mathcal{V}$), with no connections within a layer. Hidden nodes are indicated by gray filled circles and the visible nodes indicated by unfilled circles.}
  \label{fig:rbm}
\end{figure}
A common use for RBMs is to create features for use in classification. For example, binary images can be classified through a process that treats the pixel values as the visible variables $v_i$ in a RBM model [@hinton2006fast].

@@

Let $\boldsymbol x = (h_1, \dots, h_\nh, v_1,\dots,v_\nv)$ represent the states of the visible and hidden nodes in a RBM for some integers $\nv, \nh \ge 1$. Each single binary random variable, visible or hidden, will take its values in a common coding set $\mathcal{C}$, where we allow (one of) two possibilities for the coding set, $\mathcal{C}=\{0,1\}$ or $\mathcal{C}=\{-1,1\}$, with "$1$" always indicating the "high" value of the variable. While $\mathcal{C}=\{0,1\}$ may be a natural starting point, we argue in Section \ref{explorations-of-model-properties-through-simulation} that the coding $\mathcal{C}=\{-1,1\}$ induces more interpretable model properties for the RBM. A standard parametric form for probabilities corresponding to a potential vector of states, $X = (H_1, \dots, H_\nh, V_1,\dots,V_\nv)$, for the nodes is
\begin{align}
\label{eq:pmf}
f_{\boldsymbol \theta} (\boldsymbol x)\equiv P_{\boldsymbol \theta}(\boldsymbol X = \boldsymbol x) = \frac{\exp\left(\sum\limits_{i = 1}^\nv \sum\limits_{j=1}^\nh \theta_{ij} v_i h_j + \sum\limits_{i = 1}^\nv\theta_{v_i} v_i + \sum\limits_{j = 1}^\nh\theta_{h_j} h_j\right)}{\gamma(\boldsymbol \theta)}, \quad \boldsymbol x \in \mathcal{C}^{\nh + \nv} 
\end{align}
where $\boldsymbol \theta \equiv (\theta_{11}, \dots, \theta_{1\nh}, \dots, \theta_{\nv 1}, \dots, \theta_{\nv \nh}, \theta_{v_1}, \dots, \theta_{v_\nv}, \theta_{h_1}, \dots, \theta_{h_\nh}) \in \mathbb{R}^{\nv + \nh + \nv*\nh}$ denotes the vector of model parameters and the denominator 
$$
\gamma(\boldsymbol \theta) = \sum_{\boldsymbol x \in \mathcal{C}^{\nh+\nv}}\exp\left(\sum_{i = 1}^\nv \sum_{j=1}^\nh \theta_{ij} v_i h_j + \sum_{i = 1}^\nv\theta_{v_i} v_i + \sum_{j = 1}^\nh\theta_{h_j} h_j\right)
$$ 
is the normalizing function that ensures the probabilities \eqref{eq:pmf} sum to one. For $\boldsymbol x = (h_1, \dots, h_\nh, v_1, \dots, v_\nh) \in \mathcal{C}^{\nv + \nh}$ and 
\begin{align}
\boldsymbol t(\boldsymbol x) &= (h_1, \dots, h_\nh, v_1, \dots, v_\nv, v_1 h_1, \dots, v_\nv h_\nh) \in \mathcal{C}^{\nh + \nv + \nh*\nv}, \label{eq:t}
\end{align}
let $\mathcal{T} = \{\boldsymbol t(\boldsymbol x): \boldsymbol x \in \mathcal{C}^{\nv + \nh}\} \subset \mathbb{R}^{\nv + \nh + \nv * \nh}$ be the set of possible values for the vector of variables needed to compute probabilities \eqref{eq:pmf} in the model, and write $Q_{\boldsymbol \theta}(\boldsymbol x) = \sum\limits_{i = 1}^\nh\sum\limits_{j=1}^\nv \theta_{ij} h_i v_j + \sum\limits_{i = 1}^\nh\theta_{hi} h_i + \sum\limits_{j = 1}^\nv\theta_{vj} v_j$ for the "neg-potential" function. The RBM model is parameterized by $\boldsymbol \theta$ containing two types of parameters, main effects and interaction effects. The main effects parameters ($\{\theta_{v_i}, \theta_{h_j}\}_{\substack{i = 1, \dots, \nv,\\j = 1, \dots, \nh}}$) weight the values of the visible $v_i$ and hidden $h_j$ nodes in probabilities \eqref{eq:pmf} and the interaction effect parameters ($\theta_{ij}$) weight the values of the connections $v_i h_j$, or dependencies, between hidden and visible layers.

@@

## Degeneracy, instability, and uninterpretability
The highly flexible nature of a RBM (having as it does $\nh + \nv + \nh*\nv$ parameters) creates at least three kinds of potential issues of model impropriety that we will call *degeneracy*, *instability*, and *uninterpretability*. In this section we define these, consider how to quantify them in a RBM, and point out relationships among them.

@@

In Random Graph Model theory, *model degeneracy* means there is a disproportionate amount of probability placed on only a few elements of the sample space, $\mathcal{X}$, by the model [@handcock2003assessing]. For random graph models, $\mathcal{X}$ denotes all possible graphs that can be constructed from a set of nodes and an exponentially parameterized random graph model has a distribution with a probability mass function of the form 
\begin{align*}
f_{\boldsymbol \theta} (\boldsymbol x) = \frac{\exp\left(\boldsymbol \theta^T \boldsymbol t(\boldsymbol x)\right)}{\gamma(\boldsymbol \theta)}, \boldsymbol x \in \mathcal{X},
\end{align*}
where $\boldsymbol \theta \in \Theta \subset \mathbb{R}^q$ is the model parameter, and $\boldsymbol t: \mathcal{X} \rightarrow \mathbb{R}^q$ is a vector of statistics based on the adjacency matrix of a graph. Here, as earlier for RBMs, $\gamma(\boldsymbol \theta) = \sum_{\boldsymbol x \in \mathcal{X}} \exp\left(\boldsymbol \theta^T \boldsymbol t(\boldsymbol x)\right)$ is the normalizing function. Let $C$ denote the convex hull of the potential outcomes of sufficient statistics, $\{\boldsymbol t(\boldsymbol x): \boldsymbol x \in \mathcal{X}\}$, under the model above. @handcock2003assessing classifies an exponentially parameterized random graph model at $\boldsymbol \theta$ as *near-degenerate* if the mean value of the vector of sufficient statistics under $\boldsymbol \theta$, $\boldsymbol \mu(\boldsymbol \theta) = \text{E}_{\boldsymbol \theta}\boldsymbol t( \boldsymbol X)$, is close to the boundary of $C$. Intuitively, if a model is near-degenerate in the sense that only a small number of elements of the sample space $\mathcal{X}$ have positive probability, the expected value $\text{E}_{\boldsymbol \theta}\boldsymbol t( \boldsymbol X)$ is an average of that same small number of values of $t( \boldsymbol x)$ (defining the boundary of the hull $C$) and can be expected to *not* be pulled deep into the interior of $C$.

A RBM model can be thought to exhibit an analogous form of *near-degeneracy* when there is a disproportionate amount of probability placed on a small number of elements in the sample space of visibles and hiddens, $\mathcal{C}^{\nv + \nh}$. Using the idea of @handcock2003assessing, when the random vector $\boldsymbol t(\boldsymbol x) = (v_1, \dots, v_\nv, h_1, \dots, h_\nh, v_1 h_1, \dots, v_V h_\nh ) \in \mathcal{T} \equiv \{\boldsymbol t(\boldsymbol x): \boldsymbol x \in \mathcal{C}^{\nh + \nv} \}$ from \eqref{eq:t}, appearing in the neg-potential function $Q_{\boldsymbol \theta}(\cdot)$, has a mean vector $\boldsymbol \mu(\boldsymbol \theta) \in \mathbb{R}^{\nv + \nh + \nv * \nh}$ close to the boundary of the convex hull of $\mathcal{T}$, and the RBM model can be said to exhibit near-degeneracy at $\boldsymbol \theta \in \mathbb{R}^{\nv + \nh + \nh*\nv}$. Here the mean of $\boldsymbol t(\boldsymbol x)$ is
\begin{align*}
\boldsymbol \mu(\boldsymbol \theta) = \text{E}_{\boldsymbol \theta} \boldsymbol t(\boldsymbol X)  &= \sum\limits_{x \in \mathcal{C}^{\nv + \nh}} \left\{ \boldsymbol t(\boldsymbol x) f_{\boldsymbol \theta} (\boldsymbol x) \right\} \\
&= \sum\limits_{x \in \mathcal{C}^{\nv + \nh}} \left\{ \boldsymbol t(\boldsymbol x)\dfrac{\exp\left(\sum\limits_{i = 1}^\nv \sum\limits_{j=1}^\nh \theta_{ij} v_i h_j + \sum\limits_{i = 1}^\nv\theta_{v_i} v_i + \sum\limits_{j = 1}^\nh\theta_{h_j} h_j\right)}{\sum\limits_{\boldsymbol x \in \mathcal{C}^{\nh + \nv}}\exp\left(\sum\limits_{i = 1}^\nv \sum\limits_{j=1}^\nh \theta_{ij} v_i h_j + \sum\limits_{i = 1}^\nv\theta_{v_i} v_i + \sum\limits_{j = 1}^\nh\theta_{h_j} h_j\right)}\right\}.
\end{align*}

@@

In other words, if the probability ratio \eqref{eq:elpr} is too large, then a RBM model sequence will exhibit large probability shifts for very small changes in data configurations (i.e., will exhibit instability). Such instability can be a concern in a model for reasons similar to those related to degeneracy: as outcomes vary over the sample space, the geography of probabilities is extremely rugged, with deep pits following sharp mountains. Recall the applied example of RBM models as a means to classify images. For data as pixels in an image, the instability result in Proposition \ref{prp:instab1} manifests itself as a one pixel change in an image (one component of the visible vector) resulting in a large shift in the probability, which in turn could result in a vastly different classification of the image. Examples of this kind of behavior have been presented in @szegedy2013intriguing for deep learning models, in which a one pixel change in a test image results in a wildly different classification.

>>>>

---
title: "Properties and Bayesian fitting of restricted Boltzmann machines"
author: 
  - name: Andee Kaplan
    shortname: Kaplan A.,
    numaddress: 1
    corresponding: true
  - name: Daniel Nordman
    shortname: D. Nordman, and
    corresponding: false
    numaddress: 2
  - name: Stephen Vardeman
    shortname: S. Vardeman
    numaddress: "2,3"
    corresponding: false
shortauthor: "KAPLAN \\textsc{et al}"
address:
  - numaddress: 1
    orgdiv: Department of Statistical Science
    orgname: Duke University
    orgstate: North Carolina
    orgcountry: USA
  - numaddress: 2
    orgdiv: Department of Statistics
    orgname: Iowa State University
    orgstate: Iowa
    orgcountry: USA
  - numaddress: 3
    orgdiv: Department of Industrial and Manufacturing Systems Engineering
    orgname: Iowa State University
    orgstate: Iowa
    orgcountry: USA
corresponding: true
corresponding_name: Andee Kaplan
corresponding_address: Department of Statistical Science, Duke University, P.O. Box 90251, Durham, NC 27708-0251
corresponding_email: andrea.kaplan@duke.edu
output: 
  rmarkdown::pdf_document:
    keep_tex: true
    template: ../resources/latex/wileynjd-ams_article.tex
    includes:
      in_header: ../resources/latex/header.tex
    toc: no
bibliography: [../resources/latex/refs_rbm.bib]
abstract: |
   A restricted Boltzmann machine (RBM) is an undirected graphical model constructed for discrete or continuous random variables, with two layers, one hidden and one visible, and no conditional dependency within a layer. In recent years, RBMs have risen to prominence due to their connection to deep learning. By treating a hidden layer of one RBM as the visible layer in a second RBM, a deep architecture can be created. RBMs thereby are thought to have the ability to encode very complex and rich structures in data, making them attractive for supervised learning. However, the generative behavior of RBMs largely is unexplored and typical fitting methodology does not easily allow for uncertainty quantification in addition to point estimates. In this paper, we discuss the relationship between RBM parameter specification in the binary case and model properties such as degeneracy, instability and uninterpretability. We also describe the associated difficulties that can arise with likelihood-based inference and further discuss the potential Bayes fitting of such (highly flexible) models, especially as Gibbs sampling (quasi-Bayes) methods often are advocated for the RBM model structure.
keywords: Degeneracy, Instability, Classification, Deep Learning, Graphical Models
---

@@

The data mining and machine learning communities recently have shown great interest in deep learning, specifically in stacked restricted Boltzmann machines (RBMs) [see @salakhutdinov2009deep; @salakhutdinov2012efficient; @srivastava2013modeling; @le2008representational for examples]. An RBM is a probabilistic undirected graphical model (for discrete or continuous random variables) with two layers, one hidden and one visible, with no conditional dependency within a layer [@smolensky1986information]. These models reportedly have been used with success in classification of images [@larochelle2008classification; @srivastava2012multimodal]. However, the model properties are largely unexplored in the literature and the commonly cited fitting methodology remains heuristic-based and relies on rough approximation [@hinton2006fast]. Additionally, the current fitting methodology does not easily allow for interval estimation of parameters in the model, and instead only provides for point estimation. In this paper, we provide steps toward a fuller understanding of the model class and its properties from the perspective of statistical model theory, and we then explore the possibility of a rigorous fitting methodology with natural uncertainty quantification. We find the RBM model class to be concerning in two fundamental ways.

First, the models can be unsatisfactory as conceptualizations of how data are generated. That is, recalling @fisher1922mathematical, the aim of a statistical model is to represent data in a compact way. Neyman and Box further state that a model should "provide an explanation of the mechanism underlying the observed phenomena" [@lehmann1990model; @box1967discrimination]. At issue, simulation from RBMs often can produce data lacking realistic variability so that such models may thereby fail to satisfactorily reflect an observed data generation process. Such behavior relates to model degeneracy (or near degeneracy), which is a statistical concern in that, realistically, many data processes of interest are not degenerate in their spectrum of potential outcomes. For example, when sampling data from a nearly degenerate RBM used to model an imaging process, only a small set of output possibilities receives nearly all probability, and thus a sample of images will all be copies of the same one image (or small number of images). An example of ten 4-pixel images simulated from a nearly degenerate RBM model is compared to ten 4-pixel images simulated from a non-degenerate RBM model in Figure \ref{fig:sample-models}. The degenerate model places almost all probability on one outcome, causing the image to be generated repeatedly, whereas the non-degenerate model shows more potentially realistic variation. \par

@@

In addition to such degeneracy, we find that RBMs easily can exhibit types of instability, related to how sensitive the model probabilities can be to small difference in data outcomes. In practice, this may be seen when a single pixel change in an image results in a wildly different classification in an image classification problem. Occurrences of such behavior recently have been documented in RBMs [@li2014biclustering], as well as other deep architectures [@szegedy2013intriguing; @nguyen2014deep]. We describe (related) potential issues of model instability, degeneracy and uninterpretability for the RBM class (which are related properties) and examine the presence of these in Section \ref{explorations-of-model-properties-through-simulation} through simulations of small, manageable examples.

@@

Furthermore, these MCMC algorithms involve updating potentially many latent variables (hiddens), which can critically influence convergence in MCMC-based likelihood methods. Applying basic statistical principles in fitting RBM models of tractable size, we compare three fully Bayesian techniques involving MCMC, which are computationally more accessible than direct maximum likelihood and also aim to avoid parts of an RBM parameter space that yield unattractive models. As might be expected, with greater computational complexity comes an increase in fitting accuracy, but at the cost of practical feasibility.

As a factor compounding the computational challenges, issues in model fitting traceable to model flexibility potentially are more concerning. For an RBM model with enough hidden variables, any distribution for the visibles can be approximated arbitrarily well [@le2008representational; @montufar2011refinements; and @montufar2011expressive]. However, for binary data, the empirical distribution of an observed training set of visible variables provides a highest likelihood benchmark, before any parametric model class even is introduced and applied to obtain a refinement of model fit. As a consequence, we find that any fully principled fitting method based on the likelihood for an RBM with enough hidden variables will seek to reproduce the (discrete) empirical distribution of a training set. This aspect can be undesirable, and perhaps even unexpected compared to most modeling scenarios, in that no "smoothed distribution" may result when fitting an RBM model of sufficient size with a rigorous likelihood-based method. We therefore are led to be skeptical that models that involve these structures (like deep Boltzmann machines) can achieve useful prediction or inference in a principled way without intentionally limiting the flexibility of the fitted model.

@@

While applications of the RBM have been claimed to produce some successes in classification problems, it is unclear if the model class allows one to go beyond fitting to other statistical matters of importance in using the models, such as quantification of the uncertainty of estimation and the formulation of predictive distributions. These are tied closely to the probability properties of the RBM model for explaining data generation, and our exposition intends to contribute to a better understanding in this direction.

@@

A restricted Boltzmann machine (RBM) is an undirected graphical model specified for discrete or continuous random variables, binary variables most commonly being considered. In this paper, we consider the binary case for concreteness. An RBM architecture has two layers, hidden ($\mathcal{H}$) and visible ($\mathcal{V}$), with no dependency connections within a layer. An example of this structure is in Figure \ref{fig:rbm} with the hidden nodes indicated by gray circles and the visible nodes indicated by white circles.
\begin{figure}
  \centering
  \resizebox{\linewidth}{!}{\input{../resources/images/tikz/rbm.tikz}}
  \caption{An example restricted Boltzmann machine (RBM), consisting of two layers, one hidden ($\mathcal{H}$) and one visible ($\mathcal{V}$), with no connections within a layer. Hidden nodes are indicated by gray filled circles and the visible nodes indicated by unfilled circles.}
  \label{fig:rbm}
\end{figure}
A common use for RBMs is to create features for use in classification. For example, binary images can be classified through a process that treats the pixel values as the visible variables $v_i$ in an RBM model [@hinton2006fast].

@@

Let $\boldsymbol x = (h_1, \dots, h_\nh, v_1,\dots,v_\nv)$ represent the states of the visible and hidden nodes in an RBM for some integers $\nv, \nh \ge 1$. Each single binary random variable, visible or hidden, will take its values in a common coding set $\mathcal{C}$, where we allow (one of) two possibilities for the coding set, $\mathcal{C}=\{0,1\}$ or $\mathcal{C}=\{-1,1\}$, with "$1$" always indicating the "high" value of the variable. While $\mathcal{C}=\{0,1\}$ may be a natural starting point, we argue in Section \ref{explorations-of-model-properties-through-simulation} that the coding $\mathcal{C}=\{-1,1\}$ induces more interpretable model properties for the RBM. A standard parametric form for probabilities corresponding to a potential vector of states, $X = (H_1, \dots, H_\nh, V_1,\dots,V_\nv)$, for the nodes is
\begin{align}
\label{eq:pmf}
f_{\boldsymbol \theta} (\boldsymbol x)\equiv P_{\boldsymbol \theta}(\boldsymbol X = \boldsymbol x) = \frac{\exp\left(\sum\limits_{i = 1}^\nv \sum\limits_{j=1}^\nh \theta_{ij} v_i h_j + \sum\limits_{i = 1}^\nv\theta_{v_i} v_i + \sum\limits_{j = 1}^\nh\theta_{h_j} h_j\right)}{\gamma(\boldsymbol \theta)}, \quad \boldsymbol x \in \mathcal{C}^{\nh + \nv} 
\end{align}
where $\boldsymbol \theta \equiv (\theta_{11}, \dots, \theta_{1\nh}, \dots, \theta_{\nv 1}, \dots, \theta_{\nv \nh}, \theta_{v_1}, \dots, \theta_{v_\nv}, \theta_{h_1}, \dots, \theta_{h_\nh}) \in \mathbb{R}^{\nv + \nh + \nv*\nh}$ denotes the vector of model parameters and the denominator 
$$
\gamma(\boldsymbol \theta) = \sum_{\boldsymbol x \in \mathcal{C}^{\nh+\nv}}\exp\left(\sum_{i = 1}^\nv \sum_{j=1}^\nh \theta_{ij} v_i h_j + \sum_{i = 1}^\nv\theta_{v_i} v_i + \sum_{j = 1}^\nh\theta_{h_j} h_j\right)
$$ 
is the normalizing function that ensures the probabilities \eqref{eq:pmf} sum to one. For $\boldsymbol x = (h_1, \dots, h_\nh, v_1, \dots, v_\nh) \in \mathcal{C}^{\nv + \nh}$ and 
\begin{align}
\boldsymbol t(\boldsymbol x) &= (h_1, \dots, h_\nh, v_1, \dots, v_\nv, v_1 h_1, \dots, v_\nv h_\nh) \in \mathcal{C}^{\nh + \nv + \nh*\nv}, \label{eq:t}
\end{align}
let $\mathcal{T} = \{\boldsymbol t(\boldsymbol x): \boldsymbol x \in \mathcal{C}^{\nv + \nh}\} \subset \mathbb{R}^{\nv + \nh + \nv * \nh}$ be the set of possible values for the vector of variables needed to compute probabilities \eqref{eq:pmf} in the model, and write $Q_{\boldsymbol \theta}(\boldsymbol x) = \sum\limits_{i = 1}^\nh\sum\limits_{j=1}^\nv \theta_{ij} h_i v_j + \sum\limits_{i = 1}^\nh\theta_{hi} h_i + \sum\limits_{j = 1}^\nv\theta_{vj} v_j$ for the "neg-potential" function. The RBM model is parameterized by $\boldsymbol \theta$ containing two types of parameters, main effects and interaction effects. The main effects parameters ($\{\theta_{v_i}, \theta_{h_j}\}_{\substack{i = 1, \dots, \nv,\\j = 1, \dots, \nh}}$) weight the values of the visible $v_i$ and hidden $h_j$ nodes in probabilities \eqref{eq:pmf}, and the interaction effect parameters ($\theta_{ij}$) weight the values of the connections $v_i h_j$, or dependencies, between hidden and visible layers.

@@

## Degeneracy, instability, and uninterpretability
The highly flexible nature of an RBM (having as it does $\nh + \nv + \nh*\nv$ parameters) creates at least three kinds of potential issues of model impropriety that we will call *degeneracy*, *instability*, and *uninterpretability*. In this section we define these, consider how to quantify them in an RBM, and point out relationships among them.

@@

In Random Graph Model theory, *model degeneracy* means there is a disproportionate amount of probability placed on only a few elements of the sample space, $\mathcal{X}$, by the model [@handcock2003assessing]. For random graph models, $\mathcal{X}$ denotes all possible graphs that can be constructed from a set of nodes, and an exponentially parameterized random graph model has a distribution with a probability mass function of the form 
\begin{align*}
f_{\boldsymbol \theta} (\boldsymbol x) = \frac{\exp\left(\boldsymbol \theta^T \boldsymbol t(\boldsymbol x)\right)}{\gamma(\boldsymbol \theta)}, \boldsymbol x \in \mathcal{X},
\end{align*}
where $\boldsymbol \theta \in \Theta \subset \mathbb{R}^q$ is the model parameter, and $\boldsymbol t: \mathcal{X} \rightarrow \mathbb{R}^q$ is a vector of statistics based on the adjacency matrix of a graph. Here, as earlier for RBMs, $\gamma(\boldsymbol \theta) = \sum_{\boldsymbol x \in \mathcal{X}} \exp\left(\boldsymbol \theta^T \boldsymbol t(\boldsymbol x)\right)$ is the normalizing function. Let $C$ denote the convex hull of the potential outcomes of sufficient statistics, $\{\boldsymbol t(\boldsymbol x): \boldsymbol x \in \mathcal{X}\}$, under the model above. @handcock2003assessing classifies an exponentially parameterized random graph model at $\boldsymbol \theta$ as *near-degenerate* if the mean value of the vector of sufficient statistics under $\boldsymbol \theta$, $\boldsymbol \mu(\boldsymbol \theta) = \text{E}_{\boldsymbol \theta}\boldsymbol t( \boldsymbol X)$, is close to the boundary of $C$. Intuitively, if a model is near-degenerate in the sense that only a small number of elements of the sample space $\mathcal{X}$ have positive probability, the expected value $\text{E}_{\boldsymbol \theta}\boldsymbol t( \boldsymbol X)$ is an average of that same small number of values of $t( \boldsymbol x)$ (defining the boundary of the hull $C$) and can be expected to *not* be pulled deeply into the interior of $C$.

An RBM model can be thought to exhibit an analogous form of *near-degeneracy* when there is a disproportionate amount of probability placed on a small number of elements in the sample space of visibles and hiddens, $\mathcal{C}^{\nv + \nh}$. Using the idea of @handcock2003assessing, when the random vector $\boldsymbol t(\boldsymbol x) = (v_1, \dots, v_\nv, h_1, \dots, h_\nh, v_1 h_1, \dots, v_V h_\nh ) \in \mathcal{T} \equiv \{\boldsymbol t(\boldsymbol x): \boldsymbol x \in \mathcal{C}^{\nh + \nv} \}$ from \eqref{eq:t}, appearing in the neg-potential function $Q_{\boldsymbol \theta}(\cdot)$, has a mean vector $\boldsymbol \mu(\boldsymbol \theta) \in \mathbb{R}^{\nv + \nh + \nv * \nh}$ close to the boundary of the convex hull of $\mathcal{T}$, and the RBM model can be said to exhibit near-degeneracy at $\boldsymbol \theta \in \mathbb{R}^{\nv + \nh + \nh*\nv}$. Here the mean of $\boldsymbol t(\boldsymbol x)$ is
\begin{align*}
\boldsymbol \mu(\boldsymbol \theta) = \text{E}_{\boldsymbol \theta} \boldsymbol t(\boldsymbol X)  &= \sum\limits_{x \in \mathcal{C}^{\nv + \nh}} \left\{ \boldsymbol t(\boldsymbol x) f_{\boldsymbol \theta} (\boldsymbol x) \right\} \\
&= \sum\limits_{x \in \mathcal{C}^{\nv + \nh}} \left\{ \boldsymbol t(\boldsymbol x)\dfrac{\exp\left(\sum\limits_{i = 1}^\nv \sum\limits_{j=1}^\nh \theta_{ij} v_i h_j + \sum\limits_{i = 1}^\nv\theta_{v_i} v_i + \sum\limits_{j = 1}^\nh\theta_{h_j} h_j\right)}{\sum\limits_{\boldsymbol x \in \mathcal{C}^{\nh + \nv}}\exp\left(\sum\limits_{i = 1}^\nv \sum\limits_{j=1}^\nh \theta_{ij} v_i h_j + \sum\limits_{i = 1}^\nv\theta_{v_i} v_i + \sum\limits_{j = 1}^\nh\theta_{h_j} h_j\right)}\right\}.
\end{align*}

@@

In other words, if the probability ratio \eqref{eq:elpr} is too large, then an RBM model sequence will exhibit large probability shifts for very small changes in data configurations (i.e., will exhibit instability). Such instability can be a concern in a model for reasons similar to those related to degeneracy: as outcomes vary over the sample space, the geography of probabilities is extremely rugged, with deep pits following sharp mountains. Recall the applied example of RBM models as a means to classify images. For data as pixels in an image, the instability result in Proposition \ref{prp:instab1} manifests itself as a one pixel change in an image (one component of the visible vector), resulting in a large shift in the probability, which in turn could result in a vastly different classification of the image. Examples of this kind of behavior have been presented in @szegedy2013intriguing for deep learning models, in which a one pixel change in a test image results in a wildly different classification.

Additionally, S-unstable RBM model sequences may be connected formally to the near-degeneracy of Section \ref{near-degeneracy} (in which model sequences place all probability on a small portion of their sample spaces). To see this, define an arbitrary modal set of possible outcomes (i.e. set of highest probability outcomes) in RBM models with parameters $\boldsymbol \theta_\nv, \nv \ge 1$ as
$$
M_{\epsilon, \boldsymbol \theta_\nv} \equiv \left\{\boldsymbol v \in \mathcal{C}^\nv: \log P_{\boldsymbol \theta_\nv}(\boldsymbol v) > (1-\epsilon)\max\log\limits_{\boldsymbol v^*}P_{\boldsymbol \theta_\nv}(\boldsymbol v^*) + \epsilon\min\log\limits_{\boldsymbol v^*}P_{\boldsymbol \theta_\nv}(\boldsymbol v^*) \right\}
$$
for a given $0 < \epsilon < 1$. Then, S-unstable model sequences are guaranteed to be degenerate, as the following result shows.
